Baseline Drift Estimation for Air Quality Data Using Quantile Trend Filtering

### 2019-04-18 Manuscript/detrendify_aoas.tex
simulation text edits

We compare three criteria for choosing the smoothing parameter for quantile trend filtering:  $\V{\lambda}$ chosen using SIC \Eqn{SIC} (\texttt{detrendr\_SIC}); $\V{\lambda}$ chosen using the validation method with the validation set consisting of every 5th observation (\texttt{detrendr\_valid}); and $\V{\lambda}$ chosen using the proposed scaled eBIC criteria \Eqn{eBIC} (\texttt{detrendr\_eBIC}). For the second study we also examine the effect of the non-crossing quantile constraint by estimating the quantile trends separately and choosing $\V{\lambda}$ using the proposed scaled eBIC criteria (\texttt{detrendr\_Xing}). We do not include \texttt{detrendr\_Xing} in the first study because the difference in quantiles is large enough that we would not expect the non-crossing constraint to make a difference.

We also compare the performance of our quantile trend filtering method with three previously published methods: \texttt{npqw} which is the quantile-ll method described in \cite{Racine2017}, code was obtained from the author; \texttt{qsreg} in the \texttt{fields} R package and described in \cite{Oh2011} and \cite{nychka1995nonparametric}; \texttt{rqss} available in the \texttt{quantreg} package and described in \cite{KoenkerNgPortnoy1994}.  Only our detrend methods guarantee non-crossing quantiles. The regularization parameter $\V{\lambda}$ for the \texttt{rqss} method is chosen using a grid search and minimizing the SIC criteria separately for each quantile as described in \cite{KoenkerNgPortnoy1994}, the regularization parameter for \texttt{qsreg} was chosen using generalized cross-validation separately for each quantile based on the quantile criterion \cite{Oh2011}.

@@

To compare performance in estimating quantile trends in the absence of a signal component, three simulation designs from \cite{Racine2017} were considered. For all designs $t = 1, \ldots, n$,  $x(t) = t/n$, and the response $\V{y}$ was generated as
\begin{eqnarray*}
	y(t) & = & \sin(2\pi x(t)) + \epsilon(x(t))
\end{eqnarray*}
The errors were independent and had distributions:
\begin{itemize}
	\item Gaussian: $\epsilon(x(t)) \sim N\left(0, \left(\frac{1+x(t)^2}{4}\right)^2\right)$
	\item Beta: $\epsilon(x(t)) \sim Beta(1, 11-10x(t))$
	\item Mixed normal: $\epsilon(x(t))$ is simulated from a mixture of $N(-1,1)$ and  $N(1,1)$ with mixing probability $x(t)$.
\end{itemize}
The true quantile trends and an example simulated data set is show in \Fig{sim1}. One hundred datasets were generated of sizes 300, 500 and 1000.

For each method quantile trends were estimated for $\tau = \{0.05, 0.25, 0.5, 0.75, 0.95\}$. For each quantile trend and method the root mean squared error was calculated as $\mbox{RMSE} = \sqrt{\frac{1}{n}\sum_{i=1}^n (\hat{q}_{\tau}(t_i) - q_\tau(t_i))^2}$. \Fig{quantile_mse} shows the mean RMSE plus or minus twice the standard error for each method, quantile level, and sample size. In all three designs the proposed detrend methods are either better than or comparable to existing methods. Overall the \texttt{detrend\_eBIC} performs best, and especially in the mixed normal design our methods have lower RMSEs for the 5\textsuperscript{th} and 95\textsuperscript{th} quantiles. The \texttt{npqw} method performs particularly poorly in the mixed normal design due to the fact that it assumes the data come from a scale-location model, which is violated in this case.

@@

%% ----------------------------------------------------------------------
%% Peak Detection
%% ----------------------------------------------------------------------
\subsection{Peak Detection}
\label{sec:peak_detection}
%	\begin{figure}[h]
%		\caption{Example of simulated peaks, baseline, and observed measurements.}
%		\includegraphics[width = \linewidth]{Figures/ex_peaks.png}
%	\end{figure}
The second simulation design is closely motivated by our air quality analysis problem. We assume that the measured data can be represented by
\begin{eqnarray*}
	y(t_i) & = & \theta(t_i) + s(t_i) + \VE{\varepsilon}{i},
\end{eqnarray*}
where $t_i = i$ for $i = 1, ..., n$, $\theta(t)$ is the drift component that varies smoothly over time, $s(t)$ is the true signal at time $t$, and $\VE{\varepsilon}{i}$ are i.i.d.\@ errors distributed as $N(0, 0.25^2)$. We generate $\theta(t)$ using cubic natural spline basis functions with degrees of freedom sampled from a Poisson distribution with mean parameter equal to $n/100$,  and coefficients drawn from an exponential distribution with rate 1. The true signal function $s(t)$ is assumed to be zero with peaks generated using the Gaussian density function. The number of peaks is sampled from a binomial distribution with size equal to $n$ and probability equal to $0.005$ with location parameters uniformly distributed between $1$ and $n-1$ and bandwidths uniformly distributed between $2$ and $12$. The simulated peaks were multiplied by a factor that was randomly drawn from a normal distribution with mean 20 and standard deviation of 4. An example dataset with 4 signal peaks is shown in \Fig{peaks_class_eg}. One hundred datasets were generated for each $n=\{500, 1000, 2000, 4000\}$.

>>>>

We compare three criteria for choosing the smoothing parameter for quantile trend filtering:  $\V{\lambda}$ chosen using SIC \Eqn{SIC} (\texttt{detrendr\_SIC}); $\V{\lambda}$ chosen using the validation method with the validation set consisting of every 5th observation (\texttt{detrendr\_valid}); and $\V{\lambda}$ chosen using the proposed eBIC criterion \Eqn{eBIC} (\texttt{detrendr\_eBIC}). For the second study we also examine the effect of the non-crossing quantile constraint by estimating the quantile trends separately and choosing $\V{\lambda}$ using eBIC (\texttt{detrendr\_Xing}). We do not include \texttt{detrendr\_Xing} in the first study because the difference in quantiles is large enough that we would not expect the non-crossing constraint to make a difference.

We also compare the performance of our quantile trend filtering method with three previously published methods, none of which guarantee non-crossing quantiles:

\begin{itemize}
	\item \texttt{npqw}: The local linear quantile method (quantile-ll) described in \cite{Racine2017}. Code was obtained from the author.
	\item \texttt{qsreg}: Quantile smoothing splines described in \cite{Oh2004period} and \cite{nychka1995nonparametric} and implemented in the \texttt{fields} R package \citep{fields}. The regularization parameter was chosen using generalized cross-validation.
	\item \texttt{rqss}: Quantile trend filtering with $k=1$ available in the \texttt{quantreg} package and described in \cite{KoenkerNgPortnoy1994}. The regularization parameter is chosen using a grid search and minimizing the SIC \Eqn{SIC} as described in \cite{KoenkerNgPortnoy1994}.
\end{itemize}

@@

To compare performance in estimating quantile trends in the absence of a signal component, three simulation designs from \cite{Racine2017} were considered. For all designs $t = 1, \ldots, n$,  $x(t) = t/n$, and the response $\V{y}$ was generated as
\begin{eqnarray*}
	y(t) & = & \sin(2\pi x(t)) + \epsilon(x(t))
\end{eqnarray*}
The errors were simulated as independent draws from the following distributions:
\begin{itemize}
	\item Gaussian: $\epsilon(x(t)) \sim N\left(0, \left(\frac{1+x(t)^2}{4}\right)^2\right)$
	\item Beta: $\epsilon(x(t)) \sim Beta(1, 11-10x(t))$
	\item Mixed normal: $\epsilon(x(t))$ is simulated from a mixture of $N(-1,1)$ and  $N(1,1)$ with mixing probability $x(t)$.
\end{itemize}
The true quantile trends and an example simulated data set is show in \Fig{sim1}. One hundred datasets were generated of sizes 300, 500 and 1000.

Quantile trends were estimated for $\tau = \{0.05, 0.25, 0.5, 0.75, 0.95\}$ and the root mean squared error was calculated as $\mbox{RMSE}(\tau_j) = \sqrt{\frac{1}{n}\sum_{i=1}^n (\hat{\theta}_{ij} - \theta_{ij})^2}$, where $\theta_{ij}$ is the true value of the $\tau_j$th quantile of $y$ at $t=i$. \Fig{quantile_mse} shows the mean RMSE plus or minus twice the standard error for each method, quantile level, and sample size. In all three designs the proposed \texttt{detrend} methods are either better than or comparable to existing methods. Overall the \texttt{detrend\_eBIC} performs best. In the mixed normal design, specifically, our methods have lower RMSEs for the 5\textsuperscript{th} and 95\textsuperscript{th} quantiles. The \texttt{npqw} method performs particularly poorly on the mixed normal design due to the violation of the assumption that the data come from a scale-location model.

@@

%% ----------------------------------------------------------------------
%% Peak Detection
%% ----------------------------------------------------------------------
\subsection{Peak Detection}
\label{sec:peak_detection}
%	\begin{figure}[h]
%		\caption{Example of simulated peaks, baseline, and observed measurements.}
%		\includegraphics[width = \linewidth]{Figures/ex_peaks.png}
%	\end{figure}
The second simulation design is closely motivated by our air quality analysis problem. We assume that the measured data can be represented by
\begin{eqnarray*}
	y(t_i) & = & \theta(t_i) + s(t_i) + \VE{\varepsilon}{i},
\end{eqnarray*}
where $t_i = i$ for $i = 1, ..., n$, $\theta(t)$ is the drift component that varies smoothly over time, $s(t)$ is the true signal at time $t$, and $\VE{\varepsilon}{i}$ are i.i.d.\@ errors distributed as $N(0, 0.25^2)$. We generate $\theta(t)$ using cubic natural spline basis functions with degrees of freedom sampled from a Poisson distribution with mean parameter equal to $n/100$,  and coefficients drawn from an exponential distribution with rate 1. The true signal function $s(t)$ is assumed to be zero with peaks generated using the Gaussian density function. The number of peaks is sampled from a binomial distribution with size equal to $n$ and probability equal to $0.005$ with location parameters uniformly distributed between $1$ and $n-1$ and bandwidths uniformly distributed between $2$ and $12$. The simulated peaks were multiplied by a factor that was randomly drawn from a normal distribution with mean 20 and standard deviation of 4. An example dataset with 4 signal peaks is shown in \Fig{peaks_class_eg}. One hundred datasets were generated for each $n\in \{500, 1000, 2000, 4000\}$.
